---
title: "Antimicrobial Susceptibility Analysis"
output: html_document
---

# Setup

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(emmeans)
library(scales)
library(showtext)
library(svglite)
library(brglm2)

showtext_auto(TRUE)
font_add(family = "TimesNewRoman", regular = "C:/Windows/Fonts/times.ttf",
         italic  = "C:/Windows/Fonts/timesi.ttf")
font_add(family = "TimesExtraBold", regular = "C:/Windows/Fonts/timesbd.ttf")
```

-   *tidyverse* is used for general data wrangling;
-   *brglm2* is a bias-reduced variation of the binomial logistic regression model to account for datasets such as this one whereby there can be complete separation in the data, meaning outcomes could be 100% susceptible or 100% resistant;
-   *emmeans* is a method for computing probabilities from the model, and making Tukey-adjustd pairwise comparisons
-   the *scales* package enables some plot formatting
-   using *showtext* to load standard fonts, like Times New Roman
-   *svglite* for downloading plots as svg files

# Loading Data

```{r}
data_susceptibility <- read_csv("datasets/microbiology_cultures_cohort.csv") |>
  mutate(across(c(pat_enc_csn_id_coded, order_proc_id_coded), as.double))
```
Dataset contains:
- 283715 unique patients
- 55 different antibiotics
- 315 different organisms

# Filtering Organisms

```{r}
data_organisms <- data_susceptibility |>
  filter(organism %in% c("STREPTOCOCCUS PNEUMONIAE", 
                         "STAPHYLOCOCCUS AUREUS",
                         "STAPH AUREUS {MRSA}",
                         "KLEBSIELLA PNEUMONIAE",
                         "PSEUDOMONAS AERUGINOSA",
                         "ESCHERICHIA COLI"))
```
Reduced dataset contains:
- 51778 unique patients
- 50 different antibiotics
- 12 different organisms, across 5 species

# Further Wrangling
Removal of low frequency antibiotics and organisms to reduce skew
```{r}
data_wrangled <- data_organisms |>
  group_by(antibiotic) |>
  filter(n() > 100) |>  
  ungroup() |>
  group_by(organism) |>
  filter(n() > 400) |>
  ungroup()
```
Abbreviating antibiotic names, for plot readability
```{r}
abx_lookup <- c("Amoxicillin/Clavulanic Acid" = "AMC",
                "Amikacin" = "AMK",
                "Ampicillin" = "AMP",
                "Aztreonam" = "ATM",
                "Ceftazidime" = "CAZ",
                "Ciprofloxacin" = "CIP", 
                "Clindamycin" = "CLI",
                "Colistin" = "COL",
                "Ceftaroline" = "CPT",
                "Cefuroxime" = "CXM",
                "Ceftriaxone" = "CRO",
                "Cefotetan" = "CTT",
                "Cefotaxime" = "CTX",
                "Ceftazidime/Avibactam" = "CZA",
                "Cefazolin" = "CZO",
                "Ceftolozane/Tazobactam" = "CZT",
                "Daptomycin" = "DAP",
                "Doripenem" = "DOR",
                "Doxycycline" = "DOX",
                "Erythromycin" = "ERY",
                "Ertapenem" = "ETP", 
                "Cefepime" = "FEP",
                "Fosfomycin" = "FOS",
                "Cefoxitin" = "FOX",
                "Gentamicin" = "GEN", 
                "Imipenem" = "IPM",
                "Imipenem/Ebactam" = "IMR",
                "Cephalexin/Cephalothin" = "LEX/CEP",
                "Linezolid" = "LNZ",
                "Levofloxacin" = "LVX",
                "Meropenem" = "MEM", 
                "Moxifloxacin" = "MFX",
                "Nitrofurantoin" = "NIT",  
                "Oxacillin" = "OXA",
                "Penicillin" = "PEN",
                "Piperacillin" = "PIP",
                "Ampicillin/Sulbactam" = "SAM",
                "Trimethoprim/Sulfamethoxazole" = "SXT",  
                "Tetracycline" = "TCY",
                "Tigecycline" = "TGC",
                "Ticarcillin" = "TIC",
                "Tobramycin" = "TOB",
                "Piperacillin/Tazobactam" = "TZP",
                "Vancomycin" = "VAN")
```
Creating an antibiotic class variable
```{r}
class_lookup <- c("AMK" = "Aminoglycoside",
                  "GEN" = "Aminoglycoside",
                  "TOB" = "Aminoglycoside",
                  "AMC" = "BL/BLI", 
                  "CZA" = "BL/BLI", 
                  "CZT" = "BL/BLI",
                  "IMR" = "BL/BLI",
                  "SAM" = "BL/BLI",
                  "TZP" = "BL/BLI",
                  "DOR" = "Carbapenem",
                  "ETP" = "Carbapenem",
                  "IPM" = "Carbapenem", 
                  "MEM" = "Carbapenem", 
                  "CAZ" = "Cephalosporin",
                  "CPT" = "Cephalosporin",
                  "CRO" = "Cephalosporin",
                  "CTT" = "Cephalosporin",
                  "CTX" = "Cephalosporin",
                  "CXM" = "Cephalosporin",
                  "CZO" = "Cephalosporin", 
                  "FEP" = "Cephalosporin",
                  "FOX" = "Cephalosporin", 
                  "LEX/CEP" = "Cephalosporin",
                  "CIP" = "Fluoroquinolone", 
                  "LVX" = "Fluoroquinolone",
                  "MFX" = "Fluoroquinolone",
                  "SXT" = "Folate inhibitor", 
                  "VAN" = "Glycopeptide",
                  "CLI" = "Lincosamide",
                  "DAP" = "Lipopeptide",
                  "ERY" = "Macrolide",
                  "ATM" = "Monobactam",
                  "NIT" = "Nitrofuran",
                  "LNZ" = "Oxazolidinone",
                  "AMP" = "Penicillin", 
                  "OXA" = "Penicillin",
                  "PEN" = "Penicillin", 
                  "PIP" = "Penicillin", 
                  "TIC" = "Penicillin",
                  "FOS" = "Phosphonic acid",
                  "COL" = "Polymyxin",
                  "DOX" = "Tetracycline",
                  "TCY" = "Tetracycline",
                  "TGC" = "Tetracycline")
```
Recoding antibiotic abbreviations, classes, grouping subspecies by species, and recoding names into scientific format
```{r}
data_wrangled <- data_wrangled |>
  mutate(antibiotic = recode(antibiotic, !!!abx_lookup),
         class = recode(antibiotic, !!!class_lookup),
         species = recode(organism,
                          `ESCHERICHIA COLI` = "E. coli",
                          `KLEBSIELLA PNEUMONIAE` = "K. pneumoniae",
                          `PSEUDOMONAS AERUGINOSA` = "P. aeruginosa",
                          `MUCOID PSEUDOMONAS AERUGINOSA` = "P. aeruginosa",
                          `PSEUDOMONAS AERUGINOSA (NON-MUCOID CF)` = "P. aeruginosa",
                          `STAPH AUREUS {MRSA}` = "S. aureus",
                          `STAPH AUREUS(COLONY VARIANT - SMALL COLONY OR OTHER MORPHOTYPE)` = "S. aureus", 
                          `STAPHYLOCOCCUS AUREUS` = "S. aureus",
                          `STREPTOCOCCUS PNEUMONIAE` = "S. pneumoniae"),
         organism = recode(organism,
                           `ESCHERICHIA COLI` = "E. coli",
                           `KLEBSIELLA PNEUMONIAE` = "K. pneumoniae",
                           `PSEUDOMONAS AERUGINOSA` = "P. aeruginosa",
                           `MUCOID PSEUDOMONAS AERUGINOSA` = "Mucoid P. aeruginosa",
                           `PSEUDOMONAS AERUGINOSA (NON-MUCOID CF)` = 
                          "Non-mucoid P. aeruginosa",
                          `STAPH AUREUS {MRSA}` = "MRSA",
                          `STAPH AUREUS(COLONY VARIANT - SMALL COLONY OR OTHER MORPHOTYPE)` = "S. aureus morphological variant", 
                          `STAPHYLOCOCCUS AUREUS` = "S. aureus",
                          `STREPTOCOCCUS PNEUMONIAE` = "S. pneumoniae"))
```

# Plotting Susceptibility Outcomes Across Antibiotics and Species
Adding species counts for facet labels
```{r}
species_n <- data_wrangled |>
  count(species) |>
  mutate(label = paste0("italic('", species, "')~'(n = ", n, ")'"))

data_plot <- data_wrangled |>
  left_join(specoes_n, by = "species")
```
Plot!
```{r fig.width=12, fig.height=8}
data_plot |>
ggplot(aes(x = antibiotic, fill = susceptibility)) +
  geom_bar() +
  scale_y_log10(labels = trans_format("log10", math_format(10^.x))) +
  facet_wrap(~ label, labeller = label_parsed, 
             ncol = 2, 
             scales = "free_x") +
  scale_fill_manual(breaks = c("Susceptible", "Intermediate", "Resistant", 
                               "Inconclusive", "Null"),
                    values = c("Susceptible" = "#1f78b4",
                               "Intermediate" = "#FDBF00",
                               "Resistant" = "#E31A1C",
                               "Inconclusive" = "#8C564B",
                               "Null" = "grey70")) +   
  theme(axis.text.x = element_text(angle = 45, 
                                   hjust = 1, 
                                   size = 15),
        axis.title.x = element_text(face = "bold", 
                                    size = 25, 
                                    family = "TimesExtraBold"),
        axis.title.y = element_text(face = "bold", 
                                    size = 25, 
                                    family = "TimesExtraBold",
                                    margin = margin(r = 8)),
        strip.text = element_text(size = 32, 
                                  face = "italic", 
                                  family = "TimesNewRoman"),
        legend.title = element_text(family = "TimesExtraBold", 
                                    face = "bold", 
                                    size = 25),
        legend.text = element_text(family = "TimesNewRoman", 
                                   size = 20),
        legend.position = "right",
        legend.key.height = unit(1.5, "lines"),
        legend.spacing.y = unit(0.8, "lines"),
        panel.grid.major.x = element_blank(),
        panel.grid.minor = element_blank(),
        panel.grid.major.y = element_line(color = "grey85"),
        axis.ticks.length = unit(3, "pt"),
        panel.spacing = unit(2, "lines")) +
  labs(x = "Antibiotic (Abbreviated)",
       y = "Number of Isolates (log[10] scale)",
       fill = "Susceptibility Outcome")
```
# Binomial Logistic Regression Model
Converting to **susceptibility outcome** to a logical variable, dropping intermediary or null values
```{r}
all_logical <- data_wrangled |>
  filter(susceptibility %in% c("Susceptible", "Resistant")) |>
  mutate(suscep_logical = susceptibility == "Susceptible") |>
  group_by(organism, class) |>
  summarise(susceptible = sum(suscep_logical),
            resistant = sum(!suscep_logical),
            .groups = "drop")
```
Fitting model, using *brglmFit* to account for complete separation within data
```{r}
all_bin_model <- glm(cbind(susceptible, resistant) ~ organism * class,
                     data = all_logical,
                     family = binomial,
                     method = "brglmFit")
```
Model analysis
```{r}
all_emm <- emmeans(all_bin_model, ~ class | organism, type = "response")
emm_df <- as.data.frame(all_emm)
```
Computing significance from pairs
```{r}
pairwise_results <- pairs(all_emm, adjust = "tukey") |>
  as.data.frame() |>
  filter(!is.na(SE)) |>
  mutate(p.signif = case_when(p.value < 0.001 ~ "***",
                              p.value < 0.01 ~ "**",
                              p.value < 0.05 ~ "*",
                              TRUE ~ "ns"))
```
Separately ranking probabilities from highest to lowest
```{r}
ranked_classes <- emm_df |>
  group_by(organism) |>
  arrange(dplyr::desc(prob), .by_group = TRUE) |>
  filter(!is.na(prob))
```
Manually extracting relevant pairs for plotting
```{r}
signif_data <- data.frame(
  organism = c("E. coli", "E. coli", "E. coli", "E. coli", 
               "E. coli", "E. coli", "E. coli", "E. coli", 
               "E. coli", "E. coli",
               "K. pneumoniae", "K. pneumoniae", "K. pneumoniae", "K. pneumoniae", 
               "K. pneumoniae", "K. pneumoniae", "K. pneumoniae", "K. pneumoniae", 
               "K. pneumoniae", "K. pneumoniae", 
               "Mucoid P. aeruginosa",
               "Non-mucoid P. aeruginosa", "Non-mucoid P. aeruginosa", 
               "Non-mucoid P. aeruginosa", "Non-mucoid P. aeruginosa", 
               "Non-mucoid P. aeruginosa", "Non-mucoid P. aeruginosa", 
               "P. aeruginosa", "P. aeruginosa", "P. aeruginosa", "P. aeruginosa", 
               "P. aeruginosa", "P. aeruginosa", "P. aeruginosa", "P. aeruginosa",
               "S. aureus", "S. aureus", "S. aureus", "S. aureus", 
               "S. aureus", "S. aureus", "S. aureus", "S. aureus", 
               "S. aureus morphological variant", "S. aureus morphological variant",
               "S. aureus morphological variant", "S. aureus morphological variant",
               "MRSA", "MRSA", "MRSA", "MRSA", 
               "MRSA", "MRSA", "MRSA", "MRSA", 
               "S. pneumoniae", "S. pneumoniae", "S. pneumoniae", "S. pneumoniae",
               "S. pneumoniae", "S. pneumoniae"),
  class = c("Nitrofuran", "Phosphonic acid", "Aminoglycoside", "BL/BLI",
            "Monobactam", "Cephalosporin","Fluoroquinolone","Folate inhibitor",
            "Tetracycline", "Penicillin",
            "Aminoglycoside", "BL/BLI", "Fluoroquinolone", "Monobactam",
            "Cephalosporin", "Folate inhibitor", "Tetracycline", "Phosphonic acid",
            "Nitrofuran", "Penicillin",
            "Polymyxin",
            "Aminoglycoside", "Carbapenem",
            "Cephalosporin", "Fluoroquinolone",
            "Monobactam", "Penicillin",
            "Cephalosporin", "Aminoglycoside", "Carbapenem", "Fluoroquinolone",
            "Monobactam", "Polymyxin", "Penicillin", "Folate inhibitor",
            "Folate inhibitor", "Aminoglycoside", "Tetracycline", "Carbapenem",
            "Fluoroquinolone", "Lincosamide", "Penicillin", "Macrolide",
            "Penicillin", "Macrolide", "Lincosamide","Fluoroquinolone",
            "Folate inhibitor", "Aminoglycoside", "Tetracycline", "Cephalosporin",
            "Lincosamide", "Fluoroquinolone", "Macrolide", "Penicillin",
            "Cephalosporin", "Folate inhibitor", "Tetracycline", "Penicillin", 
            "Lincosamide", "Macrolide"),
  y = c(1.02, 0.98, 0.97, 0.97, 
        0.94, 0.93, 0.81, 0.74, 
        0.68, 0.57,
        1.02, 1.0, 0.97, 0.97, 
        0.97, 0.9, 0.83, 0.75, 
        0.49, 0.05,
        1.1,
        0.66, 0.72, 
        0.77, 0.63,
        0.77, 0.76,
        0.95, 0.94, 0.88, 0.85, 
        0.81, 0.79, 0.74, 0.41,
        1.02, 1.0, 0.96, 0.91,
        0.73, 0.69, 0.61, 0.59,
        0.44, 0.21, 0.32, 0.41,
        1.0, 0.97, 0.96, 0.68, 
        0.44, 0.24, 0.2, 0.07,
        0.95, 0.84, 0.82, 0.81,
        0.8, 0.74),
  label = c("***", "***", "***", "***",
            "***", "***", "***", "***",
            "***", "***",
            "***", "***", "***", "***",
            "***", "***", "***", "***",
            "***", "***",
            "ns",
            "***", "***",
            "***", "***",
            "***", "***",
            "***", "***", "***", "***", 
            "***", "***", "***", "*",
            "*", "***", "***", "*",
            "***", "***", "***", "***", 
            "***", "***", "***", "***", 
            "*", "*", "**", "***",
            "***", "***", "***", "***", 
            "*", "***", "***", "***",
            "***", "***"))
```
